## Exercise 2: nations

```{r}
source(here::here("scripts/setup.R"))
```

Let's first load a dataset:

```{r}
nations <- read_csv(file = here::here("data/nations.csv"))

nations[1:100,] %>%
   kbl(caption = "100 first rows of the 'nations' dataset") %>%
   kable_paper(full_width = FALSE) %>%
   scroll_box(width = "100%", height = "300px")
```

<p>&nbsp;</p>

These data consist of 5697 observations of 11 attributes.

Here is a description of each attribute :
   
- `iso2c` : Two-letter codes for each country, assigned by the International Organization for Standardization
- `iso3c` : Three-letter codes for each country, assigned by the International Organization for Standardization
- `country` : Country name
- `year` : Year in which the data apply
- `gdp_percap` : Gross Domestic Product per capita in current international dollars, corrected for purchasing power in different territories
- `life_expect` : Life expectancy at birth, in years
- `population` : Estimated total population at mid-year, including all residents apart from refugees
- `birth_rate` : Live births during the year per 1'000 people, based on mid-year population estimate
- `neonat_mortal_rate` : Neonatal mortality rate: babies dying before reaching 28 days of age, per 1,000 live births in a given year
- `region` : The region to which the country belongs
- `income` : The level income of the country

### Wrangling

1. From the `nations` tibble, extract a tibble `longevity` of dimension
   175 x 5 that contains the `country`, `gdp_percap`, `life_expect`,
   `population` and `region` columns.
   The dataset should be filtered for the observations in 2016 where the
   columns `life_expect` and `gdp_percap` are not NA.
   
```{r}
   ## Your code goes here
longevity <-nations %>%
   filter(year == 2016, 
          !is.na(life_expect),
          !is.na(gdp_percap)) %>% 
   select(country,
          gdp_percap, 
          life_expect,
          population, 
          region)

longevity %>% 
   kbl(caption = "Data from the year 2016") %>% 
   kable_paper(full_width = FALSE) %>% 
   scroll_box(width = "100%", height = "300px")
```
<p>&nbsp;</p>
2. From `longevity`, extract a new tibble `ea_na_75_85` of dimension 15 x 5
   that contains the `country`, `gdp_percap`, `life_expect`, `population` and
   `region` columns.
   The dataset should be filtered for countries in `"East Asia & Pacific"` or
   `"North America"`, with `life_expect` between 75 and 85 included.
   It should be sorted by decreasing `life_expect`.
      
```{r}
ea_na_75_85 <- longevity %>%
   filter(region == "East Asia & Pacific" | region == "North America", 
          life_expect >= 75, 
          life_expect <= 85) %>%
   arrange(desc(life_expect))

ea_na_75_85 %>% 
   kbl(caption = "Countries in East Asia & Pacific and North America, with life expectancy between 75 and 85") %>%  
   kable_paper(full_width = FALSE) %>% 
   scroll_box(width = "100%", height = "300px")
```
<p>&nbsp;</p>
3. From `longevity` again, extract a tibble `top_10_perc_us` of dimension
   19 x 6 that contains the `country`, `gdp_percap`, `life_expect`,
   `population`, `region` and `perc_rank` columns, where
   `perc_rank` is a new column corresponding to the percentile rank for
   `life_expect`. The dataset should be sorted
   by decreasing `perc_rank` and filtered for countries with top 10%
   `perc_rank` (i.e., `perc_rank` >= 0.9), plus `"United States"` (whose rank
   may lie outside the top 10%).
      
```{r}
   ## Your code goes here
top_10_perc_us <- longevity %>%
   mutate(perc_rank = percent_rank(life_expect)) %>% 
   arrange(desc(perc_rank)) %>% 
   filter(perc_rank >= 0.9 | country == "United States")

top_10_perc_us %>%
   kbl(caption = "Top 10% of countries in terms of life expectancy + United States") %>%  
   kable_paper(full_width = F) %>% 
   scroll_box(width = "100%", height = "300px")

```
<p>&nbsp;</p>   
4. From `nations`, extract a tibble `gdp_by_region` of dimension 189 x 3
   that contains the `region`, `year` and `total_gdp` columns, where
   `total_gdp` is a new column containing the total real GDP by
   `region` and `year`, with real GDP being the product of `gdp_percap` and
   `population`. The unit of `total_gdp` should be trillions of dollars.
      
```{r}
gdp_by_region <- nations %>%
   mutate(total_gdp = gdp_percap*population/1000000000000) %>% 
   group_by(region, year) %>%
   summarize(total_gdp = sum(total_gdp, na.rm = TRUE))

gdp_by_region %>% 
   kbl(caption = "Total GDP (in trillions of dollars) by region and by year") %>%  
   kable_paper(full_width = FALSE) %>%
   scroll_box(width = "100%", height = "300px")

```
<p>&nbsp;</p>   
5. From `nations`, extract a tibble `p_countries` of dimension 5 x 2
   that contains `income` and `p` columns. The dataset should be
   contain data from 2016 and `p` should be a new column
   containing the proportions of countries with `life_expect` over 70 by
   `income`.
      
```{r}
p_countries <- nations %>%
   group_by(income) %>%
   filter(year == 2016) %>% 
   summarize(p = sum(life_expect > 70, na.rm = TRUE)/n())
   
p_countries %>% 
   kbl(caption = "Proportion of countries with life expectancy over 70, by income classification") %>%  
   kable_paper(full_width = FALSE) %>%
   column_spec(c(1,2), width="100px")
```
     
### Visualization

6. Draw multiple horizontal boxplots for the life expectancy against region.
   To improve the readability:
       - use `fct_reorder()` to sort the column `region` by
          median of `life_expect`,
       - `coord_flip()` to make the boxplots horizontal,
       - add a red line to show the overall median life expectancy.
      
```{r}
b <- nations %>%
   ggplot(aes(x = fct_reorder(region, life_expect, median, na.rm = TRUE), y = life_expect, fill = region)) +
   geom_boxplot(show.legend = FALSE) +
   xlab("") +
   ylab("Life expectancy") +
   coord_flip() +
   ggtitle("Life expectancy by region") +
   theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
   geom_hline(yintercept = median(nations$life_expect, na.rm = TRUE), color = "red")

b 
```

7. Draw a scatterplot for `life_exp` against `gdp_percap` with a linear
   regression line (without standard errors), using a log-scale for the
   x-axis.
      
```{r}
scatterplot1 <- nations %>%
   ggplot(aes(x=gdp_percap, y=life_expect)) +
   geom_point(aes(color=region), shape=20, na.rm=TRUE) +
   geom_smooth(method="lm", se=FALSE, na.rm=TRUE) +
   scale_x_log10() +
   xlab("GPD per capita (log-scale)") +
   ylab("Life expectancy")

scatterplot1
 
```

8. To understand how life expectancy evolves over time in different region, draw 
   a lineplot for each country and a linear regression line, faceted by region.
      
```{r}
lineplot1 <- nations %>%
   ggplot(aes(x = year, y = life_expect)) +
   geom_line(aes(color=country), alpha=0.4, na.rm=TRUE, show.legend=FALSE) +
   geom_smooth(method="lm", formula="y ~ x", show.legend=FALSE, se=FALSE, na.rm=TRUE) +
   xlab("Year") +
   ylab("Life expectancy") +
   facet_wrap(~ region, nrow=2, ncol=4)

lineplot1

```
