## Exercise 1: abalone

```{r, echo=FALSE, warning=FALSE, message=FALSE}
source(here::here("scripts/setup.R"))
```

Let's first load a dataset:

```{r}
abalone <- read_csv(file = here::here("data/abalone.csv"))

abalone[1:100,] %>% 
   kbl(caption = "Abalone dataset (100 first rows)") %>%  
   kable_paper(full_width = F) %>% 
   scroll_box(width = "100%", height = "300px")
```

<p>&nbsp;</p>   
These data consist of 4177 observations of 9 attributes.

Here is a description of each attribute :
   
- `Sex nominal` : male, female and infant
- `Length` (mm) : Longest shell measurement 
- `Diameter` (mm) : perpendicular to length 
- `Height` (mm) : with meat in shell
- `Whole weight` (grams) : whole abalone
- `Shucked weight` (grams) : weight of meat
- `Viscera weight` (grams) : gut weight (after bleeding)
- `Shell weight` (grams) : after being dried
- `Rings` (integer) : +1.5 gives the age in years

### Wrangling

1. Extract a tibble `length_sex_ring` which contains the `sex`, `diameter`,
   `height` and `rings` columns. The dataset should be filtered for
   length strictly larger than 0.6 and be sorted by increasing `sex` and
   decreasing `ring`.
   
```{r}
length_sex_ring <- abalone %>%
   filter(length > 0.6) %>% 
   select(sex, diameter, height, rings) %>% 
   arrange(sex, desc(rings))

length_sex_ring %>% 
   kbl(caption = "Abalone with length strictly larger than 0.6") %>%  
   kable_paper(full_width = F) %>% 
   scroll_box(width = "100%", height = "300px")
```
<p>&nbsp;</p>
2. Extract a tibble `count_prop` of dimension 3 x 3,
   which contains the `sex`, `count` and `prop` columns.
   The dataset should be grouped by increasing `sex` with counted numbers
   of each group. Then calculate the proportion of each count.
   
```{r}
count_prop <- abalone %>%
   group_by(sex) %>%
   summarize(count = n()) %>%
   mutate(prop = count/sum(count))

count_prop %>% 
   kbl(caption = "Number and proportion per sex") %>%  
   kable_paper(full_width = F) %>%
   column_spec(c(1,2,3), width="100px")
   
```
<p>&nbsp;</p>
3. Extract a tibble `mean_max_min` of dimension 3 x 4, which contains the
   `sex`, `weight_mean`, `weight_max` and `weight_min` columns.
   The dataset should be grouped by increasing `sex` and summarized by
   finding mean, max and min of `shucked_weight`.
   
```{r}
mean_max_min <- abalone %>%
   group_by(sex) %>%
   summarize(weight_mean = mean(shucked_weight), 
             weight_max = max(shucked_weight), 
             weight_min = min(shucked_weight))

mean_max_min %>% 
   kbl(caption = "Weight's stat per sex") %>%  
   kable_paper(full_width = F)
```
<p>&nbsp;</p>
4. Extract a tibble `filter_na` of dimension 2963 x 10,
   which contains all columns. The dataset should be filtered for diameter
   equal to `NA` or strictly greater than 0.36. The column `X` should be
   renamed as `index` and you should also reorder the columns such that
   `index`, `sex`, `length`, `diameter` and `rings` come first and then
   the other columns appear in their original order.
   
```{r}
filter_na <- abalone %>%
   filter(is.na(diameter) | diameter > 0.36) %>%
   rename(index = X) %>%
   select(index,
          sex,
          length,
          diameter,
          rings,
          everything())

filter_na %>% 
   kbl(caption = "Abalone with diameter strictly greater than 0.36") %>%  
   kable_paper(full_width = F) %>% 
   scroll_box(width = "100%", height = "300px")
```

<p>&nbsp;</p>
5. Extract a tibble `transmute_abalone` of dimension 4177 x 2,
   which contains the `whole_weight_in_mg` and `water_weight_in_mg` columns,
   corresponding respectively to the whole weight and water weight in
   milligrams.
   
```{r}
transmute_abalone <- abalone %>%
   mutate(water_weight = whole_weight - (shucked_weight+viscera_weight+shell_weight)) %>%
   transmute(whole_weight_in_mg = 1000*whole_weight, 
             water_weight_in_mg = 1000*water_weight)

transmute_abalone %>% 
   kbl(caption = "Specific weight in mg ") %>% 
   kable_paper(full_width = F) %>%
   scroll_box(width = "100%", height = "300px")
```

<p>&nbsp;</p>   
6. Extract a tibble `first_1000_rank` of dimension 1000 x 3,
   which contains the `diameter`, `rings` and `rings_rank` columns, sorted
   by ascending `rings_rank`, which is a column containing the rank
   corresponding to the value of the `rings` variable.
   Then, select the three columns and rank on the `rings` and
   filter for the rows corresponding to the first 1000 `rings_rank`.
   
```{r}
first_1000_rank <- abalone %>%
   arrange(rings) %>%
   mutate(rings_rank = c(1:length(rings))) %>%
   select(diameter, 
          rings, 
          rings_rank) %>% 
   filter(rings_rank <= 1000)

first_1000_rank %>% 
   kbl(caption = "The 1000 abalone that have the least rings") %>% 
   kable_paper(full_width = F) %>% 
   scroll_box(width = "100%", height = "300px")
```

<p>&nbsp;</p>   
7. Extract a tibble `n_distinct_rings_by_sex` of dimension 3 x 2,
   which contains the `sex` and `distinct_rings` columns.
   The dataset should be grouped by `sex` and then summarized to
   count distinct rings in each group.

```{r}
n_distinct_rings_by_sex <- abalone %>%
   group_by(sex) %>% 
   summarize(distinct_rings = n_distinct(rings))

n_distinct_rings_by_sex %>% 
   kbl(caption = "Total number of distinct rings per sex") %>%  
   kable_paper(full_width = F)
```
  
### Visualization

8. Draw a density plot of `rings`, colored by `sex`.
  
```{r}
sex_names <- c(`F` = "Female",
               `I` = "Infant",
               `M` = "Male")
d <- abalone %>%
   ggplot(aes(x = rings, color = sex)) + 
   geom_density(alpha = 0.4) +
   ggtitle("Density plot of rings") +
   ylab("") +
   xlab("Rings") +
   theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
   scale_color_manual(labels = c("Female", "Infant", "Male"), values = c("red","green", "blue"))

d 
```
  
9. Draw a histogram of the `diameter` with `binwidth` as 0.05,
   filled and faceted by `sex`.
  
```{r}
h <- abalone %>%
   ggplot(aes(x = diameter, fill = sex, col = I("black"))) +
   geom_histogram(binwidth = 0.05,
                  show.legend = F) +
   facet_wrap(~ sex, labeller = as_labeller(sex_names)) +
   ylab("Number of observations") +
   xlab("Diameter") +
   ggtitle("Diameter by sex") +
   theme(plot.title = element_text(hjust = 0.5, face = "bold"))

h
```
  
10. Draw a boxplot of `diameter` against `sex`.
   The ordering of boxes from left to right should be in descending order
   of the median `diameter` per group.
  
```{r}
b <- abalone %>%
   ggplot(aes(x = fct_reorder(sex, diameter, median, .desc = TRUE), y = diameter, fill = sex)) +
   geom_boxplot(show.legend = F) +
   xlab("Sex") +
   ylab("Diameter (in mm)") +
   scale_fill_discrete(name = "Sex", labels = sex_names) +
   ggtitle("Diameter by sex") +
   theme(plot.title = element_text(hjust = 0.5, face = "bold"))

b 
```
  
11. Draw a point plot of `diameter` against `rings` with a smooth curve
   obtained using the `loess` method, and faceted by `sex`.
  
```{r}
diameter_by_rings_pp <- abalone %>%
   ggplot(aes(x = rings, y = diameter, col = sex)) +                 
   geom_point(show.legend = F, size = 0.8) + 
   geom_smooth(method = "loess", col = I("black"), se = F) +
   ylab("Diameter (in mm)") +
   xlab("Rings") +        
   facet_wrap( ~ sex, labeller = as_labeller(sex_names)) +
   ggtitle("Diameter by the number of rings") +
   theme(plot.title = element_text(hjust = 0.5, face = "bold"))

diameter_by_rings_pp
```
  
12. Draw a violin plot of `whole_weight` against `length` faceted by `sex`.

```{r}
v <- abalone %>%
   ggplot(aes(x = length, y = whole_weight, fill = sex)) +
   geom_violin(alpha = 0.8, show.legend = F) +
   ylab("Whole weight (in grams)") +
   xlab("Length") +
   ggtitle("Whole weight by length") +
   theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
   facet_wrap( ~ sex, labeller = as_labeller(sex_names)) 
   
v
```
  